## Checking for sample correlation

There are two sources of "variation" in RNA-seq experiments:
experimental (biological) and technical. Experimental variations are
changes that might result from a perturbation that we introduced on
purpose (e.g. gene knockout, differentiation), and are obviously of
interest to the researcher. Technical variations are changes
introduced as a result of our experiment that is not related to the
intended effect (e.g. differences in reagents, transfection or
differentiation efficiencies).

In most cases, we want the experimental variations to be
more consistent (and hopefully stronger) than the technical
variations. One way we can examine that is to have biological
replicates, and to correlate their outputs to ensure that samples
treated under the same experimental conditions have similar gene
expression profiles.

### Experimental setup

We performed CRISPR on a cell line, using two independent guide RNA
against a gene of interest (guideRNA1 and guideRNA2). We also have a
control guide RNA (ctrlGuide).

We generated four biological replicates (independent transfections),
and generated RNAseq libraries. They were sequenced on Illumina,
aligned to the human genome,  genes were quantified, and their counts
normalized.

### Correlation analysis

Now, we want to see if the replicates correlate with each other. One
quick way to do this is to generate a correlation plot using the R
module `corrplot`

```{r setup}
if(! requireNamespace("corrplot",quietly=TRUE)){
    install.packages("corrplot")
}

library(corrplot)
```

First we read in the data from the file

```{r readFile}
data = read.table("CRISPR_expt.txt",sep="\t",header=T,row.names=1)
```

Then we try to correlate the samples (columns in `data`). We can try
Pearson's correlation first

```{r firstPlot}
corr = cor(data,method="pearson")
corrplot(corr,tl.cex=0.8)
```

If we take a look at the output, the circles are colored by their
correlation score (see the key on the right). Most of the circles are
colored dark blue, suggesting a high correlation coefficient, but it's
not easy to see the differences

Let's try another visualization method

```{r numberPlot}
corrplot(corr,method="number",tl.cex=0.8)
```

Now you can see the values for the Pearson correlation. As you might
have guessed from the colored circles from before, they are pretty
high (most are >0.85).

Another visualization that I like to use (only because it gives an
interesting visual cue for me) is the "ellipse" method.

```{r ellipsePlot}
corrplot(corr,method="ellipse",tl.cex=0.8)
```

This is my personal preference, but I like the idea of seeing the
"shape" of the correlation. The closer to the diagonal (and smaller
the ellipse), the higher the correlation coefficient. Again, this is
just my personal preference.

Now that we have some idea of the correlation, we might be able to see
that some samples are more closely correlated with each other than
others. While you can always visually inspect them, it is also
possible to ask `corrplot` to cluster them.

Let's try to do some hierarchical clustering with this plot

```{r hclustPlot}
corrplot(corr,method="ellipse",tl.cex=0.8,order="hclust")
```

You can now see that the samples that are more similar to each other
are being grouped closer to each other. Just visually, you might be
able to see a big "cluster" of samples in the middle, and the rest on
the outside.

I'm curious to see what "clusters" are being found by R based on the
correlation. Given that we can visually see three clusters, I can ask
`corrplot` to highlight them (based on the hierarchical clustering
results) and see if it matches our visual inspection.

```{r 3clusters}
corrplot(corr,method="ellipse",tl.cex=0.8,order="hclust",addrect=3)
```

You can see that the boxes matches our visual expectation. If we now
look at the labels, we can see that the top left cluster contains all
samples from replicate 1, while the bottom right cluster contains all
samples from replicate 2. The larger middle one contains all samples
from replicate 3 and 4.

This is telling us that we might have "batch" effects, where some of the
replicates (each of which is a different transfection experiment) is
introducing variations in the data. Since we're trying to assess the
changes in expression between CRISPR and control, the variations from
the replicates might confound our analysis, and we might get fewer
"significant" genes.

In fact, if we want to see when the clusters finally separate out the
CRISPR and control samples, I had to ask for 6 clusters

```{r 6clusters}
corrplot(corr,method="ellipse",tl.cex=0.8,order="hclust",addrect=6)
```

Now you can see that the middle cluster is divided into two smaller
clusters, containing two replicates of the control experiment, and
four replicates of the CRISPR experiment.

What if we use Spearman's correlation instead of Pearson's
correlation? Pearson's correlation assumes that there is a linear
relationship between the values of two samples (so if geneA is 2x
higher in sample2 vs sample1, then geneB is expected to be 2x higher
as well). Spearman's correlation looks for a correlation in the rank
of the values

```{r spearman}
corr = cor(data,method="spearman")
corrplot(corr,method="ellipse",tl.cex=0.8,order="hclust")
```

It's not as clear where the clusters might be in this graph.

Let's try 2 clusters:

```{r spearman2clusters}
corrplot(corr,method="ellipse",tl.cex=0.8,order="hclust",addrect=2)
```

We now see that replicates 1 and 2 are in one cluster, while
replicates 3 and 4 are in another cluster. 

Let's try with 4 clusters

```{r spearman4clusters}
corrplot(corr,method="ellipse",tl.cex=0.8,order="hclust",addrect=4)
```

Now we can recapitulate what we saw previously, where samples in
replicate 1 (and 2) are more correlated to each other than to other
replicates, whereas samples in replicate 3 and 4 are more correlated
to their corresponding experiment.

### Conclusions

Based on the graph outputs, we can probably conclude that our
biological replicates 1 and 2 are showing some "batch" effects. This
means that, unless we try to correct it (and there are some
methods out there), it will make it harder for us to perform
differential expression analysis.

We can see that biological replicates 3 and 4 seem to be
clustering based on their experimental status (CRISPR vs control). It
is possible to use these two replicates for downstream analyses.

However, we should always be cautious whenever we decide to "throw
away" data. What could be the reason behind the variation that we see? 

- Is it due to initial difficulties with the protocol, given the first
2 replicates might have been the first two attempts? 
- Should we run one more replicate to ensure that we have optimized the protocol for
consistency (and thus ensure that they cluster in a way similar to
replicate 3 and 4)?
- Can we exclude the possibility that the
differences are technical variations, rather than biological in
nature?



